---
title: "Data"
format:
  html:
    code-fold: true
    code-tools: true
---

## Link to data and codes

Please click the link below to access files.

[Github classroom link](https://github.com/anly501/dsan-5000-project-Michaelx1010)

Note that due to the large size of the data, it will not be hosted on GitHub. Instead, all the data will be hosted on the GUdomain.

## Exploratory data Analysis

### Data understanding

-   Here I'm using the [U.S. Chronic Disease Indicators (CDI)](https://catalog.data.gov/dataset/u-s-chronic-disease-indicators-cdi) data

```{r, message=FALSE, warning=FALSE}
#Loading required packages
library(tidyverse)
library(ggplot2)
library(plotly)
library(ggthemes)
library(DT)
```

-   Load the cleaned data and check for different variables and data type for each of them

```{r, message=FALSE, warning=FALSE}
#Load the data
us_chronic <- read_csv("data/us_chronic.csv")
head(us_chronic)
glimpse(us_chronic)
```

-   Checking for time span of this data set

```{r}
min(us_chronic$YearStart)
max(us_chronic$YearEnd)
```

This data ranges from 2001 to 2021.

-   Checking for different data types

```{r}
unique(us_chronic$DataValueType)
```

-   Checking for location information

```{r}
loc <- unique(us_chronic$LocationDesc)
loc
```

This comprehensive dataset includes all U.S. territories, both within and outside North America.

-   Checking for different categories of Topic and Question of chronic diseases.

```{r}
types_t <- unique(us_chronic$Topic)
types_t
```

There are a total of 17 categories, encompassing Cardiovascular diseases, Cancer, diabetes, and other areas of research inquiry.

```{r}
# Loop through each unique topic
for (topic in types_t) {
  cat("Topic:", topic, "\n")
  
# Get unique questions for the current topic
  questions <- unique(us_chronic$Question[us_chronic$Topic == topic])
  
 # Print unique questions
 cat("Questions:\n", paste(questions, collapse = "\n"), "\n\n")
}
```

We obtain a list of questions for each distinct topic, which gives a more comprehensive understanding on how the data is structured around types of topics and types of questions.

```{r}
unique(us_chronic$Stratification1)
```

### Descriptive statistics and Data Visualization

-   Years Summary

```{r}
#Year summary
summary(us_chronic[, c("YearStart", "YearEnd")])
```

-   Data value summary for each distinct data types

```{r, warning=FALSE}
#Data value summary
us_chronic %>%
  group_by(DataValueType) %>%
  summarize(
    Mean = mean(DataValueAlt, na.rm = TRUE),
    Median = median(DataValueAlt, na.rm = TRUE),
    SD = sd(DataValueAlt, na.rm = TRUE),
    Min = min(DataValueAlt, na.rm = TRUE),
    Max = max(DataValueAlt, na.rm = TRUE)
  )
```

-   visualizing distribution of categorical variables

Chronic diseases type

```{r}
p1 <- ggplot(us_chronic, aes(x = Topic, fill = Topic)) +
  geom_bar() +
  labs(title = "Frequency Distribution of Chronic diseases types",
       x = "Chronic diseases",
       y = "Frequency")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_hue()
ggplotly(p1)
```

There are lots of data on Cancer, Diabetes, and Cardiovascular diseases.

Locations

```{r}
p2 <- ggplot(us_chronic, aes(x = LocationDesc, fill = LocationDesc)) +
  geom_bar() +
  labs(title = "Frequency Distribution of Locations",
       x = "Locations",
       y = "Frequency")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_hue()
ggplotly(p2)
```

We can see that the data of each state are approximately uniformly distrbuted.

Data Source

```{r}
p3 <- ggplot(us_chronic, aes(x = DataSource, fill = DataSource)) +
  geom_bar() +
  labs(title = "Frequency Distribution of Data source",
       x = "Data Sources",
       y = "Frequency")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  scale_fill_hue()
ggplotly(p3)
```

We can see that most of our data comes from The Behavioral Risk Factor Surveillance System (BRFSS) of CDC.

-   Visualizing correlations and data grouping

1.  Prevalence of diagnosed diabetes among adults aged \>= 18 years and Obesity among adults aged \>= 18 years

```{r}
#e1 <- us_chronic %>%
#  filter(Question == "Prevalence of diagnosed diabetes among adults aged >= 18 years" |
#         Question == "Obesity among adults aged >= 18 years")
#e1 <- e1 %>%
#  filter(DataValueType == "Crude Prevalence")
#e1 <- e1 %>%
#  filter(!is.na(Overall))
#glimpse(e1)
#e1 <- e1 %>%
#  pivot_wider(names_from = c(Topic), values_from = DataValueAlt)
#colnames(e1)[ncol(e1) - 1] <- "Obesity"
#e1 <- e1 %>%
#  select(YearStart, LocationAbbr, Diabetes, Obesity)
#e1 <- e1 %>%
#  pivot_wider(names_from = c(YearStart, LocationAbbr, values_from = c(Diabetes, Obesity)))
#e1 <- e1 %>% arrange(YearStart, LocationAbbr)
#e1_filled <- e1 %>%
#  group_by(YearStart, LocationAbbr) %>%
#  mutate(Diabetes = ifelse(is.na(Diabetes), mean(Diabetes, na.rm = TRUE), Diabetes)) %>%
#  ungroup()
#e1_filled <- na.omit(e1_filled)
```

```{r}
# Filter the data
e1 <- us_chronic %>%
  filter(
    Question %in% c("Prevalence of diagnosed diabetes among adults aged >= 18 years", "Obesity among adults aged >= 18 years"),
    DataValueType == "Crude Prevalence",
    !is.na(Overall)
  )

# Pivot the data
e1 <- e1 %>%
  pivot_wider(names_from = Topic, values_from = DataValueAlt)

# Rename the Obesity column
colnames(e1)[ncol(e1) - 1] <- "Obesity"

# Select relevant columns
e1 <- e1 %>%
  select(YearStart, LocationAbbr, Diabetes, Obesity)

# Group, fill NAs, and arrange the data
e1_filled <- e1 %>%
  group_by(YearStart, LocationAbbr) %>%
  mutate(Diabetes = ifelse(is.na(Diabetes), mean(Diabetes, na.rm = TRUE), Diabetes)) %>%
  ungroup() %>%
  arrange(YearStart, LocationAbbr)

# Remove rows with NA values
e1_filled <- na.omit(e1_filled)

# View the result
glimpse(e1_filled)
```

```{r}
ep1 <- ggplot(e1_filled, aes(x = Obesity, y = Diabetes, color = LocationAbbr)) +
  geom_point() +
  labs(title = "Correlation between Diabetes and Obesity",
       x = "Obesity Prevalence (%) per state",
       y = "Diabetes Prevalence (%) per state") +
  theme_minimal()

ggplotly(ep1)
```

We observe a positive correlation between the prevalence of diabetes and obesity within a specific state across multiple years.

2.  High cholesterol prevalence among adults aged \>= 18 years and Current smoking among adults aged \>= 18 years

```{r}
# Filter the data
e2 <- us_chronic %>%
  filter(
    Question %in% c("High cholesterol prevalence among adults aged >= 18 years", "Current smoking among adults aged >= 18 years"),
    DataValueType == "Crude Prevalence",
    !is.na(Overall)
  )

# Pivot the data
e2 <- e2 %>%
  pivot_wider(names_from = Topic, values_from = DataValueAlt)

# Rename the High cholestero
colnames(e2)[ncol(e2) - 1] <- "Hc"

# Select relevant columns
e2 <- e2 %>%
  select(YearStart, LocationAbbr, Hc, Tobacco)

# Group, fill NAs, and arrange the data
e2_filled <- e2 %>%
  group_by(YearStart, LocationAbbr) %>%
  mutate(Hc = ifelse(is.na(Hc), mean(Hc, na.rm = TRUE), Hc)) %>%
  ungroup() %>%
  arrange(YearStart, LocationAbbr)

# Remove rows with NA values
e2_filled <- na.omit(e2_filled)

# View the result
glimpse(e2_filled)
```

```{r}
ep2 <- ggplot(e2_filled, aes(x = Tobacco, y = Hc, color = LocationAbbr)) +
  geom_point() +
  labs(title = "Correlation between High cholesterol prevalence and Tobacco use",
       x = "Tobacco use (%) per state",
       y = "High cholesterol prevalence (%) per state") +
  theme_minimal()

ggplotly(ep2)
```

We observe a positive correlation between the prevalence of High cholesterol prevalence and Tobacco within a specific state across multiple years.

3.  Cancer of the prostate, mortality and Binge drinking frequency among adults aged \>= 18 years who binge drink

```{r}
# Filter the data
e3 <- us_chronic %>%
  filter(
    Question %in% c("Cancer of the prostate, mortality", "Binge drinking frequency among adults aged >= 18 years who binge drink"))


e3 <- us_chronic %>%
  filter(
    Question %in% c("Cancer of the prostate, mortality", "Binge drinking frequency among adults aged >= 18 years who binge drink"),
    DataValueType %in% c( "Average Annual Crude Rate","Mean"),
    !is.na(Overall)
  )

# Pivot the data
e3 <- e3 %>%
  pivot_wider(names_from = Topic, values_from = DataValueAlt)

# Rename the Prostate Cancer column
colnames(e3)[ncol(e3) - 1] <- "ProstateCancer"

# Select relevant columns
e3 <- e3 %>%
  select(YearStart, LocationAbbr, ProstateCancer, Alcohol)

# Group, fill NAs, and arrange the data
e3_filled <- e3 %>%
  group_by(YearStart, LocationAbbr) %>%
  mutate(ProstateCancer = ifelse(is.na(ProstateCancer), mean(ProstateCancer, na.rm = TRUE), ProstateCancer)) %>%
  ungroup() %>%
  arrange(YearStart, LocationAbbr)

# Remove rows with NA values
e3_filled <- na.omit(e3_filled)

# View the result
glimpse(e3_filled)
```

```{r}
ep3 <- ggplot(e3_filled, aes(x = Alcohol, y = ProstateCancer, color = LocationAbbr)) +
  geom_point() +
  labs(title = "Correlation between Binge Drinking and Prostate Cacner mortality",
       x = "Means of binge drinking per state",
       y = "Prostate Cancer mortality (%) per state") +
  theme_minimal()

ggplotly(ep3)
```

The absence of a discernible correlation suggests that the mortality rate of prostate cancer is likely influenced by factors other than binge drinking. Other significant variables may play a crucial role in determining prostate cancer mortality rates.

### Hypothesis Generation

Based on the previous data cleaning, data wrangling, and data EDA, I formulated the following hypothesis to add onto my research question:

1.  There exists a correlation between obesity and diabetes, suggesting a potential interconnections between these two health indicators.

2.  Smoking exhibits a correlation with high cholesterol levels and other cardiovascular diseases, implying a potential association between smoking habits and cardiovascular health conditions.

The forthcoming data analysis will delve into a thorough examination of the initial research question.

### Outliers detection

In this section, I will utilize the previously examined data on obesity and diabetes as an illustrative example. Subsequently, I will recycle the established code for use in other investigative analyses.

```{r, warning=FALSE, message=FALSE}
# Histogram
ggplot(e1_filled, aes(x = Diabetes)) +
  geom_histogram() +
  labs(title = "Histogram of Diabetes data", x = "Diabetes") +
  theme_minimal()

# Boxplot
ggplot(e1_filled, aes(y = Diabetes)) +
  geom_boxplot() +
  labs(title = "Boxplot of Data", y = "Diabetes") +
  theme_minimal()

# Q-Q plot
qqnorm(e1_filled$Diabetes)
qqline(e1_filled$Diabetes, col = "red")

```

The visualizations reveal that the distribution of our data closely approximates a normal distribution. The histogram highlights a concentration of values around the mean, with a noticeable tail extending beyond a value of 15. Additionally, the boxplot illustrates the majority of the data clustering within the interquartile range, while a few outliers are observed towards the upper end of the distribution. These findings suggest a generally normal distribution with a presence of notable outliers at higher values.

### EDA Reports

(All the tabular reports could be accessed through Data Understanding and Descriptive Statistics.)

The dataset under examination is a comprehensive repository encompassing chronic disease data across U.S. territories from 2001 to 2021. Comprising 1,185,676 observations and 26 variables, each entry corresponds to a specific U.S. state in a given year. The dataset encompasses 17 major categories, spanning cardiovascular diseases, diabetes, cancer, nutritional information, kidney disease, mental health, and more. Further granularity is provided by 19 distinct data types, including crude rates, age-adjusted crude rates, means, and various other metrics.

This expansive dataset poses a significant challenge due to its scale and complexity. Through exploratory data analysis (EDA), the intricacies of these diverse categories and subcategories have been elucidated. EDA has been instrumental in unraveling the nuances of prevalence rates, mortality figures, and population statistics for tobacco users, among others. The EDA process involved thorough examination of data distributions and utilization of ggplot for in-depth visualizations, facilitating a comprehensive understanding of underlying correlations.

Armed with these insights, the groundwork has been laid for a more detailed and nuanced data analysis. The multifaceted nature of the dataset, coupled with the systematic approach taken thus far, positions the research to delve into intricate patterns and relationships, paving the way for a robust and comprehensive machine learning modeling.

### Tools and softwares

For the core of this project, I will leverage R and the powerful tools provided by the tidyverse, specifically utilizing ggplot2 for creating insightful visualizations. Additionally, to enhance the depth and breadth of the project, I plan to incorporate Python into the analysis. Python libraries such as Seaborn,Plotly and Scikit-learn will be employed to further enrich the project's dimensions and offer diverse perspectives through their advanced plotting capabilities and machine learning model constructing. This multi-language approach aims to harness the strengths of both R and Python, ensuring a comprehensive and robust analysis.






