---
title: "Data Cleaning"
format:
  html:
    code-fold: true
    code-tools: true
---
```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

### Data cleaning of the [U.S. Chronic Disease Indicators (CDI)](https://catalog.data.gov/dataset/u-s-chronic-disease-indicators-cdi)

```{r}
#Load the data
us_chronic <- read_csv("data/U.S._Chronic_Disease_Indicators__CDI_.csv")
head(us_chronic)
glimpse(us_chronic)
```
- Checking for different chronic diseases types
```{r}
# Check the categories of different disease incidents
us_chronic$Topic <- as.factor(us_chronic$Topic)
category <- levels(us_chronic$Topic)
category
```

- Checking for NA columns
```{r}
# Check for NA values in each column
na_columns <- colSums(is.na(us_chronic))

# Display columns with NA values
print(na_columns)
```
As we can see, variables StratificationCategory2, Stratification2, StratificationCategory3, Stratification3, ResponseID, StratificationCategoryID2, StratificationID2, StratificationCategoryID3, StratificationID3 are completley filled with NA values. We need to drop the columns with completely NA values.

```{r}
#Drop the columns with completely NA values
us_chronic <- us_chronic[, colSums(is.na(us_chronic)) != nrow(us_chronic)]
head(us_chronic)
glimpse(us_chronic)
```

- Transforming data to tidy

Let's first look at what is inside StratificationCategory1 and Stratification1.
```{r}
#Loop through each unique category
types_s <- unique(us_chronic$StratificationCategory1)
for (i in types_s) {
  cat("Category:", i, "\n")
  
# Get unique Types for the current category
  types <- unique(us_chronic$Stratification1[us_chronic$StratificationCategory1 == i])
  
 # Print unique types
 cat("Types:\n", paste(types, collapse = "\n"), "\n\n")
}
```
In order to transform data to tidy, I need three column variables specify Race, Gender and Overall.

```{r}
us_chronic_tidy <- us_chronic %>%
  mutate(Race = ifelse(StratificationCategory1 == "Race/Ethnicity", Stratification1, NA),
         Gender = ifelse(StratificationCategory1 == "Gender", Stratification1, NA),
         Overall = ifelse(StratificationCategory1 == "Overall", Stratification1, NA)) %>%
  
  select(-StratificationCategory1, -Stratification1) %>%
  
  mutate(RaceID = ifelse(StratificationCategoryID1 == "RACE", StratificationID1, NA),
         GenderID = ifelse(StratificationCategoryID1 == "GENDER", StratificationID1, NA),
         OverallID = ifelse(StratificationCategoryID1 == "OVERALL", StratificationID1, NA)) %>%
  
  select(-StratificationCategoryID1, -StratificationID1)

# View the resulting tidy dataset
head(us_chronic_tidy)
glimpse(us_chronic_tidy)
```
- Save the cleaned data

```{r}
write.csv(us_chronic_tidy,file='data/us_chronic.csv', row.names = FALSE) 
```

